\section{Background}
\subsection{Process Calculi}

Process calculi , sometimes referred to as process algebras are a family of languages and models for describing concurrent systems. They allow for the description of communication and synchronization between two or more concurrent processes. The algebraic laws which govern process calculi allow the process descriptions they provide to be reasoned about easily.
All process calculi allow for the following operations \cite{wiki:pa}:
\begin{itemize}
    \item Communication
    \item Sequential Composition
    \item Parallel Composition
    \item Reduction Semantics
    \item Hiding 
    \item Recursion and Replication
    \item The Null Process
\end{itemize}
\subsubsection{Communication}
Processes are able to send messages between each other. Process calculi will generally have a pair of operators defining both input and output. Formally these are often $\bar{x}\langle y \rangle$ for a process sending out message $y$ on channel $x$, and $x \left( v \right) $ for a process receiving a message on channel $x$ and binding the variable $v$ to the value of that message in subsequent processes. 
It is the type of data that can be sent/received by processes which sets apart different process calculi
\subsubsection{Sequential Composition}
Processes can potentially perform communications in order. This is signified by the sequential composition operator, often "$.$" . A process may need to wait for input on channel $x$ before continuing with other processes, which could be formally written $x\left( v \right) .P$
\subsubsection{Parallel Composition}
Processes can perform actions concurrently and independently. Process $P$ and $Q$ running in parallel, written $P|Q$ are able to communicate across any shared channels, however they are not limited to one channel only. These channels may be either synchronous, where the sending process must wait until the message is received, or asynchronous, where no such waiting is required.

\subsubsection{Reduction Semantics}
The details of reduction semantics are different for each process calculus, but the theory is the same. The process $\bar{x}\langle y \rangle . P | x \left( v \right) . Q $ reduces to the process $P|Q\left[ \frac{y}{v} \right]$, which is to say the following: the left hand process sends out message $y$ on channel $x$ and becomes the process $P$, and the right hand process receives a message ( $y$ ) on channel $x$, binding that message to the variable $v$ for the remaining processes
in $Q$.
\subsubsection{Hiding}
The ability to hide a name in a process is vital for the control of communications made in parallel. Hiding the name $x$ in $P$ could be written $P \backslash \left[ x \right]$.
\subsubsection{Recursion and Replication}
Recursion and replication allow for a process to continue indefinitely. Recursion of a process is a sequential concept and would be written $P = P.P$. Replication is the concurrent equivalent i.e. $!P = P | !P$
\subsubsection{The Null Process}
Finally, the null process, generally represented as $0$ or $\emptyset$, does not interact with any other processes. It acts as the terminal process, and is the basis for processes which actually do things.


\subsection{$\pi$-calculus and the Calculus of Communicating systems}

The applied $\pi$-calculus \cite{af01} is an extension of $\pi$-calculus \cite{mpw92} which itself as an extension of the work Robert Milner did on the Calculus of Communicating Systems (CCS) \cite{m82}. All three languages are process modelling languages, that is to say that they are used to describe concurrent processes and interactions between them. 
CCS is able to describe communications between two participants, and has all of the basic process algebra components as above. 
$\pi$-calculus provides an important extension allowing channel names to be passed along channels. This allows it to model concurrent processes whose configurations are not constant.

\subsection{Interpretation}

Trying to interpret a process based language presents several difficulties from the offset. Such an interpreter needs to be able to generate processes, switch contexts, and perform cross-channel communication very quickly as these operations, which are normally considered computationally intensive, form the basis of any process calculus. \cite{pt97} 
As such, it may be necessary either to reduce the feature set of the language in order to ensure that the interpreter performs acceptably. 

\subsection{The applied $\pi$-calculus}

As mentioned before, the applied $\pi$-calculus is based on $\pi$-calculus, but it is designed specifically to model security protocols \cite{rs13}. It is extended to include a large set of complex primitives and functions.
\subsubsection{Syntax}

The language assumes an infinite set of names and variables and a signature $\sigma$ which is the finite set of functions and their corresponding arities\cite{af01}. A function with arity 0 is considered a constant. Given these, the set of terms is described by the following grammar:
\begin{table}[hc!]
    \begin{tabular}{r l}
        $L,M,N,T,U,V ::=$ & terms \\
        $a,b,c,...,s$ & names\\
        $x,y,z$ & variables\\
        $g(M_{1},M_{2},...M_{l})$ & function application \\
    \end{tabular}
\end{table}

The type system (or sort system) comprises a set of base types such as \emph{Integer} and \emph{Key}, but also a universal \emph{Data}type. Names and variables can have any type. 
Processes have the following grammar:

\begin{table}[hc!]
    \begin{tabular}{r l}
         $P,Q,R ::=$ & processes  \\
         $\emptyset$& null process\\
        $P|Q$& parallel composition\\
        $P.Q$& sequential composition\\
        $!P$& replication \\
        $vn.P$& new \\
        $if M=N then P else Q $& conditional \\
        $u(x).P$& input \\
        $\bar{u}\langle N \rangle .P$& output \\
    \end{tabular}
\end{table} 
Where conditional acts as expected and "new" restricts the name n in p.Processes are extended as follows with active substitutions.

\begin{table}[hc!]
    \begin{tabular}{r l}
         $A,B,C ::=$ & extended processes  \\
        $P$& plain process\\
        $A|B$& process composition \\
        $vn.A$& new name \\
        $vx.A$& new variable \\
        $\left[ \frac{M}{x} \right]$&  active substitution \\
    \end{tabular}
\end{table} 

The active substitution $\left[ \frac{M}{x} \right]$ represents the process that has output $M$ before and this value is now reference-able by the name $x$.

\subsubsection{Simplified Syntax}

As the Pict language did when creating an implementation of pure $\pi$-calculus we must first simplify the syntax of the language we are using \cite{pt97}. Function application will remain the same, and the set of variables and names shall in theory still be infinite. We will do away with the null process, and assume that a process without a sequential process is implicitly followed by the null process.


\begin{table}[hc!]
    \begin{tabular}{l l}
        $\emptyset$&\verb!0!\\
        $P|Q$& \verb!P | Q!\\
        $P.Q$& \verb!P ; Q!\\
        $!P$ & \verb?!P?\\
        $vn$& \verb!new x!\\
        $if M=N then P else Q $&\verb!if p(M) then P else Q! \\
        $u(x)$&\verb!in(u,x)! \\
        $\bar{u}\langle N \rangle .P$&\verb!out(u,N)! \\
        $\left[ \frac{M}{x} \right]$&\verb!let X = M in P!\\
    \end{tabular}
\end{table} 

This will be the syntax we refer to from now on, and which we will be attempting to interpret.

\subsubsection{Starting Restrictions}

The first build of our interpreter will only be able to handle a few basic types and functions. We will also only be concerning ourselves with HTTP traffic, and to begin with our channels will be for the most part duplex handles built from sockets \cite{wiki:sockets} handled by the Haskell Network library \cite{hack:network}. The reasoning here is that getting to grips with the entirety of the low level C socket API (which is exposed by the Network.Socket module) would not be a productive use of time with respect to our main intended purpose (i.e. to handle HTTP traffic).
By limiting our initial approach, we can familiarise ourselves with some of the concepts of
sockets and later on we can expand to more generic sockets if necessary and once our language implementation is sound.

\subsection{Haskell}

Haskell is a pure non-strict functional programming language based on the $\lambda$-calculus. It is a strongly static typed language making it easy to ensure correctness of programs. It is highly expressive, but this combined with its laziness comes at a potential price in terms of execution time. We may well find that it simply is not possible to build a responsive enough system using Haskell, but there are several advantages to using it to build an interpreter.

\subsubsection{Data Type}
Haskell makes it trivial to create data types. As such we can easily use Haskell to build an abstract representation of our language, which we will later generate during the parsing process.

A small subset of this is presented below, we will go into more detail in Section \ref{sec:types}
    
\inputminted[linenos,frame=lines]{hs}{pi-calc.hs}

We can test this as follows with the basic input from our introduction (of course these currently hold no intrinsic meaning, but this will be implemented later)

\begin{code}
    ghci>(In "a" "x") `Seq' (Out "a" (TVar "x"))
    in(a,x);
    out(a,x)
\end{code}

It is also good to note that malformed structures will fail:

\begin{code}
    ghci>(In "a" "x") `Seq' (Out "a" )

    <interactive>:34:21:
        Couldn't match expected type `PiProcess'
        with actual type `Term -> PiProcess'
        In the return type of a call of `Out'
        Probable cause: `Out' is applied to too few arguments
        In the second argument of `Seq', namely `(Out "a")'
        In the expression: (In "a" "x") `Seq' (Out "a")
\end{code}

\subsubsection{Parsec}

Parsec is a monadic parser combinator library for Haskell which is fast, robust, simple and well-documented \cite{lm01}. We use parsec by building a series of low-level parsers and combining them into a single high level one.
For example, if we start with a low-level parser to match brackets, from that we can build a higher level parser which can then return the contents of those brackets as a list of strings:

\begin{minted}[linenos,frame=lines]{hs}
import Text.ParserCombinators.Parsec

openB :: Parser Char
openB = char '(' 

closeB :: Parser Char
closeB = char ')' 

betweenB :: Parser [String]
betweenB = do{
            openB;
            [out] <- endBy line closeB;
            return out;
        } 
    where
        line = sepBy word (char ',')
        word = many ( noneOf ",)")
\end{minted}

We can test this parser using the parseTest function as follows:

\begin{code}
    ghci>parseTest betweenB "(sometext,somemoretext)"
    ["sometext","somemoretext"]
\end{code}

This function also fails on malformed input

\begin{code}
    ghci>parseTest betweenB "(sometextsomemoretext"
    parse error at (line 1, column 22):
    unexpected end of input
    expecting "," or ")"
\end{code}

From here, it is only a short step to build a parser for the basic \verb!in,out! syntax of the language. The following is a slightly untidy, but quickly built proof of concept parser to demonstrate the ease with which this can be achieved:

\begin{minted}[linenos,frame=lines]{hs}
inOut :: Parser PiProcess
inOut = do {
        piIn;
        bContents <- betweenB;
        case bContents of
            [chan,message] -> return $ In chan message
            _ -> error (e "in(chan,message)")
        } <|> do {
        piOut;
        bContents <- betweenB;
        case bContents of
            [chan, message] -> return $ Out chan (TVar message)
            _ -> error (e "out(chan,message)")
        }
        where
            piIn= string "in"
            piOut=string "out" 
            e x= "malformed input " ++ x ++ " expected"
\end{minted}

And a demonstration of this in action (removing our custom show instance, as otherwise, because we chose to make it appear like the input, it seems like nothing is happening). Once more, malformed input throws errors depending on what exactly went wrong: 

\begin{code}
    ghci>parseTest inOut "in(a,b)"
    In "a" "b"
    ghci>parseTest inOut "out(a,b)"
    Out "a" b
    ghci>parseTest inOut "ot(a,b)"
    parse error at (line 1, column 1):
    unexpected "t"
    expecting "out"
    ghci>parseTest inOut "Out(a,b)"
    parse error at (line 1, column 1):
    unexpected "O"
    expecting "in" or "out"
    ghci>parseTest inOut "out( a,b)"
    Out " a" b
    ghci>parseTest inOut "out( a ,      b)"
    Out " a "       b
    ghci>parseTest inOut "out  a ,      b)"
    parse error at (line 1, column 4):
    unexpected " "
    expecting "("
    ghci>parseTest inOut "out(a,b,c)"
    *** Exception: malformed input out(chan,message) expected
\end{code}

\subsubsection{Sockets and the Network module}

Once we have parsed our input and built our syntax tree, we must then semantically analyse the input and generate the processes required. As mentioned before we will be using the Haskell Network module  for our initial implementation. This provides a high-level interface to network functionality in Haskell. We create handles from incoming connections on sockets on the server side, and on the client side we generate a handle by connecting to a foreign host on a specified port. An example
of this is as follows: 

\inputminted[linenos,frame=lines]{hs}{sockets.hs}

This small program creates two processes (which here are two GHC threads created with $forkIO$) . The client continuously reads from stdin and then sends the information over the socket, and the server receives the data from the socket and sends it back. In other words, written in our version of the applied $\pi$-calculus it does:
\begin{verbatim}
    !(in(stdin,x);
    out(socket(0.0.0.0,9000),x);
    in(socket(0.0.0.0,9000),y);
    out(stdout,y)) 
    |
    !(in(socket(0.0.0.0,9000),x);
    out(socket(0.0.0.0,9000),x))
\end{verbatim}

Which is something akin to our basic program from our introduction.

